\huge A Sphere Model for Atrial Fibrillation

### 2017-02-23 Literature Review/Atrial Fibrillation Poster FINAL.tex
final tingz

\textbf{Atrial Fibrillation} (AF) is a type of cardiac arrhythmia and one of the major causes of stroke and heart failure. AF is also the most widespread cardiac condition, with over 30 million people worldwide suffering from it. In the UK, expenses related to this disease account for more than 1 percent of the total budget of the National Health Service, which is more than 800 million pounds.\\*
 Atrial fibrillation usually manifests itself in patients with abnormal cardiac rhythm lasting short periods of time.\\

@@

\begin{columns}
\begin{column}{.54\textwidth}
	\begin{itemize}
	\item \textbf{\textit{Heartbeat Propagation}}: Each heartbeat originates as an electric signal in the sinoatrial (SA) node that propagates first into the atria, then through the atrioventricular (AV) node, through Purkinje fibres, into the lower reaches of the heart (Figure \ref{fig:heart}). The electric signal is conducted in the cardiac muscle cells thanks to polarisation (change in voltage) of the cell membrane \cite{Comtois}. 
	\end{itemize}
\end{column}

@@

	\item \textbf{ \textit{Cells Excitation Rules}}: As the electric impulse propagates, each cardiac single muscle cell can be in one of three stages:

		\begin{itemize}
 		 \item \textit{Rest/Excitable state}: the muscle cell is at rest with a negative built up potential.

  		\item \textit{Excited state}: the muscle cell is excited by the excited cells it is connected to and its voltage is at a maximum. 
  		\item \textit{Refractory state}: the muscle cell goes through a phase during which it can't be excited again, a decrease in voltage until its rest state voltage---this period is called \textit{refractory period}. The refractory period duration is modulated by the cell excitation rate and this relationship is called \textit{restitution}.
		\end{itemize}

@@

\begin{columns}
\begin{column}{.43\textwidth}
\begin{figure}
	\includegraphics[width=0.95\textwidth]{spiralbreak2tachy}
	\caption{\label{fig:fib}Idealised diagram of: (a), normal heatbeat propagation; (b), tachycardia and (c), fibrillation. }
	%\end{wrapfigure}
	\end{figure}

\end{column}
\begin{column}{.54\textwidth}
\begin{itemize}
	\item \textbf{\textit{How AF Occurs}}: AF is correlated to the amount of \textbf{ fibrosis}---which is the scarring of cells. This creates dysfunctional/inactive cells, decreasing the number of connections between the cells in cardiac tissue. This promotes a process called \textit{reentry}, during which, the electric signal wavefront propagation breaks \cite{Alonso}. Even though many treatments have been developed, the reentry mechanism is not completely understood and, as such, AF remains a major topic in medical research. \\
	\end{itemize}

@@

\begin{enumerate}
\item Replicate the model of a 2D cardiac atrium from Christensen \emph{et al.}
\item  Translate the model onto a sphere in order to simulate the effects of a more realistic morphology. 
\item Incorporate restitution of heart cells into the original model (such that the \textit{refractory period} of cells is modulated).
\item Observe the risk of AF in the new models and compare with the original model.
\end{enumerate}

@@

\end{column}
\hspace{-6em}
\begin{column}{.3\textwidth}
\caption{The structure of the sphere model upon which electric impulse propagation was implemented using a network. The connections in red are differentiated from the connection in blue in order to introduce fibrosis to the model. An effect of fibrosis is modelled by removing a percentage of the red connections. }

@@

In research by Fedotov, a model of the atrium was constructed on the surface of a triangulated sphere to model AF on a closed, heterogeneous surface \cite{Fedotov}. Inspired by this research, we built a new model which translates Christensen \emph{et al.}'s model of a two dimensional atrium onto a sphere.

@@

%\centering
\begin{figure}
\includegraphics[width=0.8\linewidth]{combchristensenspherefib}
\caption{A comparison between Christensen \emph{et al.}'s model  (a, b and c)  and the sphere model (for d, e and f ) evolving in time in a fibrotic atrium. Normal wavefront propagation is seen starting in the leftmost figures .In figures b,e the excitation wavefront propagates unevenly due to fibrosis, initiating reentry. On the right, AF has spontaneously arisen from both models, generating fibrillation.}
\end{figure}

@@

In Christensen \emph{et al.}'s model \cite{Christensen}, the refractory period of each muscle cell was fixed to be constant.
In a real heart, the refractory period depends on the heart cell activation rate. The relationship between refractory period and heart beat rate is called restitution. We introduced a linear restitution function in  Christensen \emph{et al.}'s model in order to study its effects.
\end{block}

@@

%\centering
\begin{figure}
\includegraphics[width=0.65\linewidth]{xriskcurvesphere}
\caption{Risk of AF (defined as the fraction of time spent in fibrillation) for the three models studied (Christensen /emph{et al.}'s model, restitution curve model and sphere model) averaged over 50 simulations, against the percentage of transversal connections ($\nu$) in the models. A healthy heart is expected to have $\nu$ close to 1 while a fibrotic heart has low $\nu$.}
\end{figure}

@@

A simplified mechanism of AF had been achieved by Christensen \emph{et al.'s} model through the modelling of the atrium muscle tissue while incorporating the effects of fibrosis. We implemented this model by adding more features: a more realistic heart morphology, using a sphere, and restitution. Our results show a higher risk of AF for the new models created.

@@

%\nocite{*} % Insert publications even if they are not cited in the poster
%\small{\bibliographystyle{unsrt}
%\bibliography{sample1}}
\bibitem{Comtois}
P. Comtois, J. Kneller, S. Nattel,
\emph{Of circles and spirals: Bridging the gap between the leading circle and spiral wave concepts of cardiac reentry}
Europace. 2005 Sep;7 Suppl 2:10-20.
DOI: http://dx.doi.org/10.1016/j.eupc.2005.05.011

>>>>

\textbf{Atrial Fibrillation} (AF) is a type of cardiac arrhythmia and one of the major causes of stroke and heart failure. AF is also the most widespread cardiac condition, with over 30 million people suffering from it worldwide. In the UK, expenses related to this disease account for more than 1 percent of the total National Health Service budget, costing more than 800 million pounds.\\*
 Atrial fibrillation usually manifests itself in patients with tachycardia, an abnormal hastening of cardiac beating.\\

@@

\begin{columns}
\begin{column}{.54\textwidth}
	\begin{itemize}
	\item \textbf{\textit{Heartbeat Propagation}}: Each heartbeat originates as an electric signal in the sinoatrial (SA) node that propagates first into the atria, then through the atrioventricular (AV) node, through Purkinje fibres, into the lower reaches of the heart (Figure \ref{fig:heart}). The electric signal is conducted through the cardiac muscle cells due to polarisation (a change in voltage of the cell membrane) which contracts the cell \cite{Comtois}. 
	\end{itemize}
\end{column}

@@

	\item \textbf{ \textit{Cells Excitation Rules}}: As the electric impulse propagates, each cardiac muscle cell can be in one of three stages:

		\begin{itemize}
 		 \item \textit{Rest/Excitable state}: the muscle cell is at rest with constant negative potential.								
  		\item \textit{Excited state}: the muscle cell is excited by adjacent excited cells it is connected to, making the voltage is at a maximum. 
  		\item \textit{Refractory state}: the muscle cell goes through a phase during which it can't be excited again---a steady decrease in voltage until its matches the rest state voltage---this period is called \textit{refractory period}. The refractory period duration is modulated by the cell excitation rate; this relationship is called \textit{restitution}.
		\end{itemize}

@@

\begin{columns}
\begin{column}{.43\textwidth}
\begin{figure}
	\includegraphics[width=0.95\textwidth]{spiralbreak2tachy}
	\caption{\label{fig:fib}Idealised diagram of: (a), normal heartbeat propagation; (b), tachycardia and (c), fibrillation. }
	%\end{wrapfigure}
	\end{figure}

\end{column}
\begin{column}{.54\textwidth}
\begin{itemize}
	\item \textbf{\textit{How AF Occurs}}: AF is correlated to the amount of \textbf{ fibrosis}---the scarring of cells. This creates dysfunctional/inactive cells, decreasing the number of connections between cells in the cardiac tissue. This promotes a process called \textit{reentry}, during which, the electric signal wavefront propagation chaotically breaks, initiating fibrillation \cite{Alonso}. Even though many treatments have been developed, the reentry mechanism is not completely understood and, as such, AF remains a major topic in medical research. \\
	\end{itemize}

@@

\begin{enumerate}
\item Replicate the model of the 2D cardiac atrium from Christensen \emph{et al.}
\item  Translate the model onto a sphere in order to simulate the effects of a more realistic morphology. 
\item Incorporate restitution of heart cells into the original model (such that the \textit{refractory period} of cells is modulated).
\item Observe the risk of AF in the new models and compare with the original model.
\end{enumerate}

@@

\end{column}
\hspace{-6em}
\begin{column}{.3\textwidth}
\caption{The structure of the sphere model. The connections in red (muscle fibres) are differentiated from the connection in blue (connections between fibres) in order to introduce fibrosis to the model. The effect of fibrosis is modelled by removing a percentage of the blue connections. Fibre growth from the SA node is seen.}

@@

In research by Fedotov, a model of the atrium was constructed on the surface of a triangulated sphere to model AF on a closed, heterogeneous surface \cite{Fedotov}. Inspired by this, we built a new model which translates Christensen \emph{et al.}'s model of a two dimensional atrium onto a sphere.

@@

%\centering
\begin{figure}
\includegraphics[width=0.8\linewidth]{combchristensenspherefib}
\caption{A comparison between Christensen \emph{et al.}'s model  (a, b and c)  and the sphere model (for d, e and f ) evolving in time in a fibrotic atrium. Normal wavefront propagation is seen starting in the leftmost figures .In figures b,e the excitation wavefront propagates unevenly due to fibrosis, initiating reentry. On the right, AF has spontaneously arisen from both models.}
\end{figure}

@@

In Christensen \emph{et al.}'s model, the refractory period of each muscle cell was fixed.
In a real heart, the refractory period depends on the heart cell activation rate. The relationship between refractory period and cell excitation rate is called restitution. We introduced a linear restitution function into our replica of Christensen \emph{et al.}'s model in order to study its effects.
\end{block}

@@

%\centering
\begin{figure}
\includegraphics[width=0.65\linewidth]{xriskcurvesphere}
\caption{Risk of AF (defined as the fraction of time spent in fibrillation) for the three models studied: Christensen \emph{et al.}'s model ("Kishan's Data"); our original model; a restitution model and the sphere model. Each point is an average risk over 50 simulations, using a specified percentage of transversal connections ($\nu$), mimicking fibrosis levels. A healthy heart is expected to have $\nu$ close to 1 while a fibrotic heart has low $\nu$. Incorporating restitution into original model increases the risk. The sphere model incorporating fibre growth increases the risk further, producing a more realistic model as less fibrosis is necessary to induce AF. }
\end{figure}

@@

A simplified mechanism of AF had been achieved by replication of Christensen \emph{et al.'s} model: simulating a sheet of atrial muscle tissue and incorporating the effects of fibrosis. We incorporated more features into this model: a more realistic heart morphology,with muscle fibre growth, using a sphere, and restitution. Our results show a higher risk of AF for the new models created. This produced more realistic models as less fibrosis is necessary to induce AF.
\end{block}

@@

%\nocite{*} % Insert publications even if they are not cited in the poster
%\small{\bibliographystyle{unsrt}
%\bibliography{sample1}}
\small
\bibitem{Comtois}
P. Comtois, J. Kneller, S. Nattel,
\emph{Of circles and spirals: Bridging the gap between the leading circle and spiral wave concepts of cardiac reentry}
Europace. 2005 Sep;7 Suppl 2:10-20.
DOI: http://dx.doi.org/10.1016/j.eupc.2005.05.011
